Detecting diversifying selection for a trait from within and between-species genotypes and phenotypes

### 2024-07-03 manuscript/main-first-submission.tex
Revision round 2 manuscript files.

\usepackage{biorxiv}
\usepackage[backend=biber,eprint=false,isbn=false,url=false,giveninits=true, style=apa,date=year]{biblatex}
\DeclareDelimFormat{nameyeardelim}{\addcomma\space}
\addbibresource{references.bib}

@@

\newcommand{\defEqual}{\stackrel{\text{def}}{=}}
\newcommand{\Multiply}{\cdot}
\newcommand{\MultiplyMatrix}{\times}
\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\newcommand{\tr}{^{\intercal}}
\newcommand{\inv}{^{-1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\text{Var}}
\DeclareMathOperator{\Cov}{\text{Cov}}
\newcommand{\Qst}{Q$_\text{ST}$}
\newcommand{\Fst}{F$_\text{ST}$}
\newcommand{\QstFst}{\Qst--\Fst}
\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\dnds}{\omega}
\newcommand{\pn}{\pi_N}
\newcommand{\ps}{\pi_S}
\newcommand{\pnps}{\pn / \ps}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\Spi}{i}
\newcommand{\Spj}{j}
\newcommand{\NbrTaxa}{n}
\newcommand{\Time}{t}
\newcommand{\Trait}{P}
\newcommand{\Heritability}{h^2}
\newcommand{\HeritabilitySpi}{\Heritability_{\Spi}}
\newcommand{\MeanTrait}{\bar{\Trait_{\Time}}}
\newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
\newcommand{\Root}{0}
\newcommand{\RootTrait}{\Trait_{\Root}}
\newcommand{\VarPhy}{\Var \left[\MeanTrait\right]}
\newcommand{\VecZero}{\UniDimArray{0}}
\newcommand{\VecOne}{\UniDimArray{1}}
\newcommand{\Distance}{\BiDimArray{D}}
\newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
\newcommand{\MutationRate}{\mu}
\newcommand{\SubRate}{q}
\newcommand{\NbrLoci}{L}
\newcommand{\VarPhenotype}{V_{\Trait}}
\newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
\newcommand{\VarGenetic}{V_{\mathrm{A}}}
\newcommand{\VarGeneticSpi}{V_{\mathrm{A}, \Spi}}
\newcommand{\VarEnv}{V_{\mathrm{E}}}
\newcommand{\VarMutation}{V_{\mathrm{M}}}
\newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
\newcommand{\RateBetween}{\sigma^2_{\mathrm{B}}}
\newcommand{\RateWhithin}{\sigma^2_{\mathrm{W}}}
\newcommand{\RateWhithinSpi}{\sigma^2_{\mathrm{W}, \Spi}}
\newcommand{\VecRateWhithin}{\UniDimArray{\RateWhithin}}
\newcommand{\EstRateBetween}{\widehat{\RateBetween}}
\newcommand{\EstRateWhithin}{\widehat{\RateWhithin}}
\newcommand{\NI}{\rho}
\newcommand{\EstNI}{\widehat{\rho}}
\newcommand{\StdSelection}{\sigma}
\newcommand{\VarSelection}{\StdSelection^2}

@@

Determining whether a trait is under a particular regime of selection has been a long-standing goal in evolutionary biology.
Fundamentally, distinguishing neutral evolution from selection requires determining which selective regime is supported by the observed variation of traits or sequences.
The variation of phenotypes (traits) and genotypes (sequences) can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
All these systems require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\parencite{lynch_genetics_1998, walsh_evolution_2018}.

Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait) and under stabilizing selection, while the loci affecting those traits are mostly pleiotropic (many traits associated with a given locus) with additive effects~\parencite{simons_population_2018, sella_thinking_2019}. 
Across several populations, by contrasting both trait and genetic differentiation, \QstFst\ methods have been used to determine the selective regime and to quantify the strength of selection acting on a trait~\parencite{merila_comparison_2001, leinonen_comparative_2008}.
A trait differentiation (\Qst) higher than genetic differentiation (\Fst) is interpreted as a signature of diversifying selection due to adaptation in different optimum trait value in the different populations~\parencite{lamy_qst_2012}.
Contrarily, \Qst\ lower than \Fst\ is interpreted as a signature of stabilizing selection.
However, \QstFst\ methods have been found to require many populations~\parencite{ohara_bias_2005}, and that various factors can generate a spurious signal of selection~\parencite{pujol_are_2008, edelaar_comparisons_2011}.
Moreover, the test for diversifying selection is limited to recent local adaptation since the test is based on the variation observed within a single species.
To disentangle selection from neutral evolution, trait variation can also be observed at a larger time scale.
For example, change in mean trait value accumulates linearly with time of divergence from a sister species, and also proportionally to the trait variance~\parencite{lande_genetic_1980, turelli_heritable_1984}. % #I would add some examples here that do not use expression data, and then make a clear transition to the use of expression data
Empirically, this effect can be observed for genes with larger within-species variation in gene expression level, which exhibits a faster accumulation of divergence in mean expression level~\parencite{khaitovich_neutral_2004}.
Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\parencite{walsh_evolution_2018}.
% For example, in the context of the evolution of gene expression, studies have shown that stabilizing selection is largely dominant~\parencite{whitehead_neutral_2006, gilad_natural_2006}.
% However, selective regime acting on gene expression is still controversial, and neutral evolution is still debated~\parencite{signor_evolution_2018, price_detecting_2022}. % are these studies using Gst/Fst on gene expression? Otherwise, we should move it elsewhere or add a clear transition between the sentences

To disentangle neutral evolution and selection, trait evolution can also be observed at a larger time scale.
For example, change in mean trait value accumulates linearly with time of divergence from a sister species, and also proportionally to the trait variance~\parencite{lande_genetic_1980, turelli_heritable_1984}. % #I would add some examples here that do not use expression data, and then make a clear transition to the use of expression data
Gene expression exhibits a similar accumulation as divergence in expression accumulates faster for genes with large within-species variation~\parencite{khaitovich_neutral_2004}.
Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\parencite{walsh_evolution_2018}.

% Phylogenetics (mean trait evolution)
Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\parencite{felsenstein_phylogenies_1985}.
Under neutral evolution, the change in mean trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\parencite{hansen_translating_1996}.
As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
Reconstructing the trait variation along the whole phylogeny as a Brownian process can thus constitute a null model of neutral trait evolution.
Deviations from the assumptions of the Brownian process are however well known.
When trait variation is constraint because of optimum mean trait values across or between species, the pattern of evolution can be modeled by the Ornstein-Uhlenbeck processes, which is often interpreted as a signature of stabilizing selection~\parencite{catalan_drift_2019}.
Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction without fixed optimum) is interpreted as a signature of directional selection at the phylogenetic scale~\parencite{silvestro_early_2019}.
However, studies have shown that such comparative approaches are subject to different biases~\parencite{harmon_phylogenetic_2018}.
First, a trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\parencite{hansen_translating_1996}.
In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
Second, even for a trait evolving under a neutral regime, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Those limitations, altogether with the use of mean trait estimates leaving out the variance in traits between individuals, easily generate misclassification of selection from methods at the phylogenetic scale.

% At the frontier Phylogenetics/Population-genetics
At the frontier between micro and macroevolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation together with changes in mean trait value to either describe measurement errors~\parencite{lynch_methods_1991, hansen_interpreting_2012}, incorporate values for individuals~\parencite{felsenstein_comparative_2008} or to scale the rate of change in mean trait value~\parencite{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
Within-species variation has also been used to infer diversifying selection by estimating the ratio of between to within variation of many traits and test for deviation from the average ratio across traits~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
Here, our goal was again to use both variances between and within species to determine the selective regime of a quantitative trait.
We build a novel framework that integrates trait variation at the phylogenetic and population scales together with estimates of molecular divergence at both scales.
It allowed us to define an expected ratio of normalized variance between and within species while setting the threshold of this ratio for neutral, stabilizing, and diversifying selection.
The ratio that we propose can be considered as a neutrality index for a any quantitative trait articulating trait and nucleotide variation within and between species.
Importantly, our neutrality index also leverages nucleotide divergence and polymorphism to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between species).
From the field of population genetics, our study can be seen as the macro-evolutionary generalization of \QstFst\ methods to account for phylogenetic relationships between species.
From the field of phylogenetics, our study can be seen as an alternative to the EVE model~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} for a single trait, where we set a threshold for neutral evolution by leveraging species nucleotide polymorphism and divergence.
% Basically, we are testing whether the variance of means is greater than the mean of variances

@@

For a given trait, the genetic architecture is mainly defined by the number of loci encoding the trait ($\NbrLoci$) and the random additive effect of a mutation on the trait ($a$).
New mutations are generating trait variance and the average effect of a mutation on the trait is $\RateMut = \NbrLoci \Multiply  \E [a^2]$.
At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRate$) and $\RateMut$:
\begin{gather}
    \VarMutation = 2 \MutationRate \Multiply \RateMut \label{eq:var-mutation}.
\end{gather}

While in an infinitesimal model mutations supply new genetic variants, random genetic drift depletes standing variation~\parencite{turelli_commentary_2017, barton_infinitesimal_2017, sella_thinking_2019}.
For a neutral trait at equilibrium between mutation and drift~\parencite{lynch_mutation_1998}, the additive genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$):
\begin{align}
    \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
    & = 4 \Ne \Multiply \MutationRate \Multiply \RateMut \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
\end{align}

For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is measured as the number of mutations segregating in the population divided by the length of the region.
Any segregating mutations will eventually reach fixation or extinction due to random genetic drift and $\pi$ is also at a balance between mutations and drift.
As shown in \textcite{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$):
\begin{gather}
    \pi = 4 \Ne \Multiply \MutationRate \label{eq:genetic-diversity}.
\end{gather}

We define $\RateWhithin$ as the ratio of additive genetic variance of the trait ($\VarGenetic$) over $\pi$ of any neutral genomic region of interest.
This ratio allows removing the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RateWhithin$ as a proxy of $\RateMut$:
\begin{align}
    \RateWhithin & \defEqual \frac{\VarGenetic }{\pi}, \\
    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \RateMut}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
    & = \RateMut. \label{eq:rate-pop}
\end{align}

The additive genetic variance is also equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by narrow-sense heritability ($\Heritability$; \parencite{hill_data_2008}), which leads to $\RateWhithin$ being a function of $\VarPhenotype$ and $\Heritability$:
\begin{align}
    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
\end{align}

@@

For a given species, we denote by $\MeanTrait$ the mean value of the trait across the individuals in the species at generation $\Time$.
If the trait is neutral and encoded by many loci as assumed by the infinitesimal model, $\MeanTrait$ evolves as a Brownian process~\parencite{felsenstein_phylogenies_1985, hansen_translating_1996}.
The variance of $\MeanTrait$ after $\Time$ generations, $\VarPhy$ is given \parencite{hansen_translating_1996} by:
\begin{align}
    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
    & = 4 \Time \Multiply \MutationRate \Multiply \RateMut \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
\end{align}

Moreover, for any genomic region under neutral evolution, some mutations will eventually reach fixation due to random genetic drift, resulting in a substitution of a nucleotide at the species level.
The probability of fixation ($\pfix$) of a neutral mutation is $1/2\Ne$ \parencite{kimura_probability_1962}.
We can derive the substitution rate per generation $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each newly arisen mutations $\pfix$~\parencite{mccandlish_modeling_2014}, giving:
\begin{align}
    \SubRate & = 2 \Ne \Multiply \MutationRate \Multiply \pfix, \\
    & = 2 \Ne  \Multiply \MutationRate  \Multiply \dfrac{1}{2\Ne}, \\
    & = \MutationRate. \label{eq:substitution-rate}
\end{align}
That is, if mutations are neutral, the rate of substitution within a genomic region equals the rate at which new mutations arise per generation for the same genomic region~\parencite{kimura_evolutionary_1968}.

After $\Time$ generations and assuming that no multiple substitutions occurred at the same site, the nucleotide divergence $d$, which is the fraction of the genomic region that generated a substitution, will be $\Time$ multiplied by the nucleotide substitution rate per generation ($\SubRate$):
\begin{align}
    d & = \Time \Multiply \SubRate \\
    & = \Time \Multiply \MutationRate \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
\end{align}
We define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by the nucleotide divergence of any neutral genomic region ($d$).
This ratio allows removing the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trade, giving $\RateBetween$ as another proxy of $\RateMut$:
\begin{align}
    \RateBetween & \defEqual \frac{\VarPhy}{4 d},\\
    & = \frac{4 \Time \Multiply \MutationRate \Multiply \RateMut}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
    & = \RateMut \label{eq:rate-phy}.
\end{align}

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/fig-input-output}
    \caption{
        Between species, the change along the phylogeny of the mean phenotypic trait allows the estimation of between-species trait variation, $\EstRateBetween$, which is normalized by nucleotide divergence.
        Within species, the genetic variance allows the estimation of within-species trait variation, $\EstRateWhithin$, which is  normalized by nucleotide diversity.
        $\EstNI$ is the ratio of $\EstRateBetween$ over $\EstRateWhithin$.
        Under neutral evolution, $\EstNI$ is expected to be equal to one.
        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\EstNI$ greater than one.
        Under stabilizing selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow for discarding the confounding effect on mutation rate diversity, population size and divergence time.
    }
    \label{fig:methods}
\end{figure*}

@@

\subsection*{Acknowledgements}
 We gratefully acknowledge the help of Nicolas Lartillot, Philippe Veber, Isabela Jeronimo do Ó, Anna Marcionetti, Julien Clavel, and Daniele Silvestro for their insightful discussions and Julien Joseph for his advice and reviews concerning this manuscript.

\subsection*{Competing interests:}
The authors declare no conflicts of interest.

>>>>

\usepackage{biorxiv}

@@

\newcommand{\defEqual}{\stackrel{\text{def}}{=}}
\newcommand{\Multiply}{\cdot}
\newcommand{\MultiplyMatrix}{\times}
\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\newcommand{\tr}{^{\intercal}}
\newcommand{\inv}{^{-1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\text{Var}}
\DeclareMathOperator{\Cov}{\text{Cov}}
\newcommand{\Qst}{Q$_\text{ST}$}
\newcommand{\Fst}{F$_\text{ST}$}
\newcommand{\QstFst}{\Qst--\Fst}
\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\dnds}{\omega}
\newcommand{\pn}{\pi_N}
\newcommand{\ps}{\pi_S}
\newcommand{\pnps}{\pn / \ps}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\Spi}{i}
\newcommand{\Spj}{j}
\newcommand{\NbrTaxa}{n}
\newcommand{\Time}{t}
\newcommand{\Trait}{P}
\newcommand{\Heritability}{h^2}
\newcommand{\HeritabilitySpi}{\Heritability_{\Spi}}
\newcommand{\MeanTrait}{\bar{\Trait_{\Time}}}
\newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
\newcommand{\Root}{0}
\newcommand{\RootTrait}{\Trait_{\Root}}
\newcommand{\VarPhy}{\Var \left[\MeanTrait\right]}
\newcommand{\VecZero}{\UniDimArray{0}}
\newcommand{\VecOne}{\UniDimArray{1}}
\newcommand{\Distance}{\BiDimArray{D}}
\newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
\newcommand{\MutationRatePheno}{\mu}
\newcommand{\SubRate}{q}
\newcommand{\NbrLoci}{L}
\newcommand{\VarPhenotype}{V_{\Trait}}
\newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
\newcommand{\VarGenetic}{V_{\mathrm{A}}}
\newcommand{\VarGeneticSpi}{V_{\mathrm{A}, \Spi}}
\newcommand{\VarEnv}{V_{\mathrm{E}}}
\newcommand{\VarMutation}{V_{\mathrm{M}}}
\newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
\newcommand{\RateBetween}{\sigma^2_{\mathrm{B}}}
\newcommand{\RateWhithin}{\sigma^2_{\mathrm{W}}}
\newcommand{\RateWhithinSpi}{\sigma^2_{\mathrm{W}, \Spi}}
\newcommand{\VecRateWhithin}{\UniDimArray{\RateWhithin}}
\newcommand{\EstRateBetween}{\widehat{\RateBetween}}
\newcommand{\EstRateWhithin}{\widehat{\RateWhithin}}
\newcommand{\NI}{\rho}
\newcommand{\EstNI}{\widehat{\rho}}
\newcommand{\StdSelection}{\sigma}
\newcommand{\VarSelection}{\StdSelection^2}

@@

Determining whether a trait is under a particular regime of selection has been a long-standing goal in evolutionary biology.
Fundamentally, distinguishing neutral evolution from selection requires determining which selective regime is supported by the observed variation of traits or sequences.
The variation of phenotypes (traits) and genotypes (sequences) can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
All these systems require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\citep{lynch_genetics_1998, walsh_evolution_2018}.

Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait) and under stabilizing selection, while the loci affecting those traits are mostly pleiotropic (many traits associated with a given locus) with additive effects~\citep{simons_population_2018, sella_thinking_2019}. 
Across several populations, by contrasting both trait and genetic differentiation, \QstFst\ methods have been used to determine the selective regime and to quantify the strength of selection acting on a trait~\citep{merila_comparison_2001, leinonen_comparative_2008}.
A trait differentiation (\Qst) higher than genetic differentiation (\Fst) is interpreted as a signature of diversifying selection due to adaptation in different optimum trait value in the different populations~\citep{lamy_qst_2012}.
Contrarily, \Qst\ lower than \Fst\ is interpreted as a signature of stabilizing selection.
However, \QstFst\ methods have been found to require many populations~\citep{ohara_bias_2005}, and that various factors can generate a spurious signal of selection~\citep{pujol_are_2008, edelaar_comparisons_2011}.
Moreover, the test for diversifying selection is limited to recent local adaptation since the test is based on the variation observed within a single species.
To disentangle selection from neutral evolution, trait variation can also be observed at a larger time scale.
For example, change in mean trait value accumulates linearly with time of divergence from a sister species, and also proportionally to the trait variance~\citep{lande_genetic_1980, turelli_heritable_1984}. % #I would add some examples here that do not use expression data, and then make a clear transition to the use of expression data
Empirically, this effect can be observed for genes with larger within-species variation in gene expression level, which exhibits a faster accumulation of divergence in mean expression level~\citep{khaitovich_neutral_2004}.
Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\citep{walsh_evolution_2018}.
% For example, in the context of the evolution of gene expression, studies have shown that stabilizing selection is largely dominant~\citep{whitehead_neutral_2006, gilad_natural_2006}.
% However, selective regime acting on gene expression is still controversial, and neutral evolution is still debated~\citep{signor_evolution_2018, price_detecting_2022}. % are these studies using Gst/Fst on gene expression? Otherwise, we should move it elsewhere or add a clear transition between the sentences

To disentangle neutral evolution and selection, trait evolution can also be observed at a larger time scale.
For example, change in mean trait value accumulates linearly with time of divergence from a sister species, and also proportionally to the trait variance~\citep{lande_genetic_1980, turelli_heritable_1984}. % #I would add some examples here that do not use expression data, and then make a clear transition to the use of expression data
Gene expression exhibits a similar accumulation as divergence in expression accumulates faster for genes with large within-species variation~\citep{khaitovich_neutral_2004}.
Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\citep{walsh_evolution_2018}.

% Phylogenetics (mean trait evolution)
Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\citep{felsenstein_phylogenies_1985}.
Under neutral evolution, the change in mean trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\citep{hansen_translating_1996}.
As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\citep{hansen_translating_1996, harmon_phylogenetic_2018}.
Reconstructing the trait variation along the whole phylogeny as a Brownian process can thus constitute a null model of neutral trait evolution.
Deviations from the assumptions of the Brownian process are however well known.
When trait variation is constraint because of optimum mean trait values across or between species, the pattern of evolution can be modeled by the Ornstein-Uhlenbeck processes, which is often interpreted as a signature of stabilizing selection~\citep{catalan_drift_2019}.
Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction without fixed optimum) is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
However, studies have shown that such comparative approaches are subject to different biases~\citep{harmon_phylogenetic_2018}.
First, a trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\citep{hansen_translating_1996}.
In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
Second, even for a trait evolving under a neutral regime, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Those limitations, altogether with the use of mean trait estimates leaving out the variance in traits between individuals, easily generate misclassification of selection from methods at the phylogenetic scale.

% At the frontier Phylogenetics/Population-genetics
At the frontier between micro and macroevolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation together with changes in mean trait value to either describe measurement errors~\citep{lynch_methods_1991, hansen_interpreting_2012}, incorporate values for individuals~\citep{felsenstein_comparative_2008} or to scale the rate of change in mean trait value~\citep{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
Within-species variation has also been used to infer diversifying selection by estimating the ratio of between to within variation of many traits and test for deviation from the average ratio across traits~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
Here, our goal was again to use both variances between and within species to determine the selective regime of a quantitative trait.
We build a novel framework that integrates trait variation at the phylogenetic and population scales together with estimates of molecular divergence at both scales.
It allowed us to define an expected ratio of normalized variance between and within species while setting the threshold of this ratio for neutral, stabilizing, and diversifying selection.
The ratio that we propose can be considered as a neutrality index for a any quantitative trait articulating trait and nucleotide variation within and between species.
Importantly, our neutrality index also leverages nucleotide divergence and polymorphism to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between species).
From the field of population genetics, our study can be seen as the macro-evolutionary generalization of \QstFst\ methods to account for phylogenetic relationships between species.
From the field of phylogenetics, our study can be seen as an alternative to the EVE model~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} for a single trait, where we set a threshold for neutral evolution by leveraging species nucleotide polymorphism and divergence.
% Basically, we are testing whether the variance of means is greater than the mean of variances

@@

For a given trait, the genetic architecture is mainly defined by the number of loci encoding the trait ($\NbrLoci$) and the random additive effect of a mutation on the trait ($a$).
New mutations are generating trait variance and the average effect of a mutation on the trait is $\RateMut = \NbrLoci \Multiply  \E [a^2]$.
At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
As shown in \citet{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRatePheno$) and $\RateMut$:
\begin{gather}
    \VarMutation = 2 \MutationRatePheno \Multiply \RateMut \label{eq:var-mutation}.
\end{gather}

While in an infinitesimal model mutations supply new genetic variants, random genetic drift depletes standing variation~\citep{turelli_commentary_2017, barton_infinitesimal_2017, sella_thinking_2019}.
For a neutral trait at equilibrium between mutation and drift~\citep{lynch_mutation_1998}, the additive genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$):
\begin{align}
    \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
    & = 4 \Ne \Multiply \MutationRatePheno \Multiply \RateMut \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
\end{align}

For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is measured as the number of mutations segregating in the population divided by the length of the region.
Any segregating mutations will eventually reach fixation or extinction due to random genetic drift and $\pi$ is also at a balance between mutations and drift.
As shown in \citet{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRatePheno$) and the effective population size ($\Ne$):
\begin{gather}
    \pi = 4 \Ne \Multiply \MutationRatePheno \label{eq:genetic-diversity}.
\end{gather}

We define $\RateWhithin$ as the ratio of additive genetic variance of the trait ($\VarGenetic$) over $\pi$ of any neutral genomic region of interest.
This ratio allows removing the effect of $\Ne$ and $\MutationRatePheno$, which are parameters not related to the genetic architecture of the trait, giving $\RateWhithin$ as a proxy of $\RateMut$:
\begin{align}
    \RateWhithin & \defEqual \frac{\VarGenetic }{\pi}, \\
    & = \frac{4 \Ne \Multiply \MutationRatePheno \Multiply \RateMut}{4 \Ne \Multiply \MutationRatePheno} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
    & = \RateMut. \label{eq:rate-pop}
\end{align}

The additive genetic variance is also equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by narrow-sense heritability ($\Heritability$; \citep{hill_data_2008}), which leads to $\RateWhithin$ being a function of $\VarPhenotype$ and $\Heritability$:
\begin{align}
    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
\end{align}

@@

For a given species, we denote by $\MeanTrait$ the mean value of the trait across the individuals in the species at generation $\Time$.
If the trait is neutral and encoded by many loci as assumed by the infinitesimal model, $\MeanTrait$ evolves as a Brownian process~\citep{felsenstein_phylogenies_1985, hansen_translating_1996}.
The variance of $\MeanTrait$ after $\Time$ generations, $\VarPhy$ is given \citep{hansen_translating_1996} by:
\begin{align}
    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
    & = 4 \Time \Multiply \MutationRatePheno \Multiply \RateMut \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
\end{align}

Moreover, for any genomic region under neutral evolution, some mutations will eventually reach fixation due to random genetic drift, resulting in a substitution of a nucleotide at the species level.
The probability of fixation ($\pfix$) of a neutral mutation is $1/2\Ne$ \citep{kimura_probability_1962}.
We can derive the substitution rate per generation $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRatePheno$) multiplied by the probability of fixation for each newly arisen mutations $\pfix$~\citep{mccandlish_modeling_2014}, giving:
\begin{align}
    \SubRate & = 2 \Ne \Multiply \MutationRatePheno \Multiply \pfix, \\
    & = 2 \Ne  \Multiply \MutationRatePheno  \Multiply \dfrac{1}{2\Ne}, \\
    & = \MutationRatePheno. \label{eq:substitution-rate}
\end{align}
That is, if mutations are neutral, the rate of substitution within a genomic region equals the rate at which new mutations arise per generation for the same genomic region~\citep{kimura_evolutionary_1968}.

After $\Time$ generations and assuming that no multiple substitutions occurred at the same site, the nucleotide divergence $d$, which is the fraction of the genomic region that generated a substitution, will be $\Time$ multiplied by the nucleotide substitution rate per generation ($\SubRate$):
\begin{align}
    d & = \Time \Multiply \SubRate \\
    & = \Time \Multiply \MutationRatePheno \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
\end{align}
We define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by the nucleotide divergence of any neutral genomic region ($d$).
This ratio allows removing the effect of $\Time$ and $\MutationRatePheno$, which are parameters not related to the genetic architecture of the trade, giving $\RateBetween$ as another proxy of $\RateMut$:
\begin{align}
    \RateBetween & \defEqual \frac{\VarPhy}{4 d},\\
    & = \frac{4 \Time \Multiply \MutationRatePheno \Multiply \RateMut}{4 \Time \Multiply \MutationRatePheno}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
    & = \RateMut \label{eq:rate-phy}.
\end{align}

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {figure1}
    \caption{
        Between species, the change along the phylogeny of the mean phenotypic trait allows the estimation of between-species trait variation, $\EstRateBetween$, which is normalized by nucleotide divergence.
        Within species, the genetic variance allows the estimation of within-species trait variation, $\EstRateWhithin$, which is  normalized by nucleotide diversity.
        $\EstNI$ is the ratio of $\EstRateBetween$ over $\EstRateWhithin$.
        Under neutral evolution, $\EstNI$ is expected to be equal to one.
        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\EstNI$ greater than one.
        Under stabilizing selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow for discarding the confounding effect on mutation rate diversity, population size and divergence time.
    }
    \label{fig:methods}
\end{figure*}

@@

%TC:ignore
\bibliographystyle{natbib}
\bibliography{references_bibtex}
